---
title: 'Credit Risk Modelling: Logistic Regression vs XGBoost'
author: "Bridget Murugi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```r
{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
  )
```markdown
# Introduction

This project develops and compares Logistic Regression and XGBoost models to predict credit default risk using the German Credit dataset.

The focus is on predictive performance and business-relevant trade-offs between detecting defaulters and approving good clients.
---
# Data Loading
```{r}
url <- "https://raw.githubusercontent.com/selva86/datasets/master/GermanCredit.csv"
data <- read.csv(url, stringsAsFactors = FALSE)

head(data)
```

---

# Target Variable Correction

Ensure correct labeling:

* 1 = Default (Bad)
* 0 = Non-default (Good)

```{r}
table(data$credit_risk)

data$credit_risk <- 1 - data$credit_risk

table(data$credit_risk)
```

---

# Data Preprocessing

```{r}
# Convert characters to factors
data[] <- lapply(data, function(x) {
  if (is.character(x)) as.factor(x) else x
})

library(caret)

# One-hot encoding
dummies <- dummyVars(credit_risk ~ ., data = data)
data_encoded <- predict(dummies, newdata = data)
data_encoded <- as.data.frame(data_encoded)

data_encoded$credit_risk <- data$credit_risk
```

---

# Train-Test Split

```{r}
set.seed(123)

train_index <- sample(1:nrow(data_encoded), 0.7 * nrow(data_encoded))

train <- data_encoded[train_index, ]
test  <- data_encoded[-train_index, ]
```

---

# Logistic Regression Model

```{r}
log_model <- glm(credit_risk ~ ., data = train, family = binomial)

log_probs <- predict(log_model, test, type = "response")

log_pred_05 <- ifelse(log_probs > 0.5, 1, 0)
log_pred_04 <- ifelse(log_probs > 0.4, 1, 0)
log_pred_03 <- ifelse(log_probs > 0.3, 1, 0)
```

---

# Logistic Regression Evaluation

```{r}
library(caret)

cm_log_05 <- confusionMatrix(as.factor(log_pred_05), as.factor(test$credit_risk), positive = "1")
cm_log_04 <- confusionMatrix(as.factor(log_pred_04), as.factor(test$credit_risk), positive = "1")
cm_log_03 <- confusionMatrix(as.factor(log_pred_03), as.factor(test$credit_risk), positive = "1")

cm_log_05
cm_log_04
cm_log_03
```

**Interpretation:**

* 0.5: Conservative (higher specificity)
* 0.4: Best balance
* 0.3: High recall, more false positives

---

# XGBoost Model

```{r}
library(xgboost)

train_matrix <- as.matrix(train[, -ncol(train)])
test_matrix  <- as.matrix(test[, -ncol(test)])

dtrain <- xgb.DMatrix(data = train_matrix, label = train$credit_risk)
dtest  <- xgb.DMatrix(data = test_matrix)

params <- list(
  objective = "binary:logistic",
  eval_metric = "auc",
  max_depth = 3,
  eta = 0.05,
  subsample = 0.8,
  colsample_bytree = 0.8
)

xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 150
)

xgb_probs <- predict(xgb_model, dtest)

xgb_pred_05 <- ifelse(xgb_probs > 0.5, 1, 0)
xgb_pred_04 <- ifelse(xgb_probs > 0.4, 1, 0)
xgb_pred_03 <- ifelse(xgb_probs > 0.3, 1, 0)
```

---

# XGBoost Evaluation

```{r}
cm_xgb_05 <- confusionMatrix(as.factor(xgb_pred_05), as.factor(test$credit_risk), positive = "1")
cm_xgb_04 <- confusionMatrix(as.factor(xgb_pred_04), as.factor(test$credit_risk), positive = "1")
cm_xgb_03 <- confusionMatrix(as.factor(xgb_pred_03), as.factor(test$credit_risk), positive = "1")

cm_xgb_05
cm_xgb_04
cm_xgb_03
```

---

# ROC Curve and AUC

```{r}
library(pROC)

roc_log <- roc(test$credit_risk, log_probs)
roc_xgb <- roc(test$credit_risk, xgb_probs)

auc_log <- auc(roc_log)
auc_xgb <- auc(roc_xgb)

auc_log
auc_xgb
```

---

# ROC Curve Visualization

```{r}
plot(roc_log, col = "blue", main = "ROC Curve: Logistic vs XGBoost")
plot(roc_xgb, col = "red", add = TRUE)

legend("bottomright",
       legend = c("Logistic Regression", "XGBoost"),
       col = c("blue", "red"),
       lwd = 2)
```

---

# Final Comparison

* Logistic Regression AUC ≈ 0.77
* XGBoost AUC ≈ 0.77

Threshold = 0.4 provides the best trade-off between sensitivity and specificity.

---

# Business Interpretation

* False Negatives → financial loss
* False Positives → lost revenue

Threshold tuning is critical in aligning model outputs with business objectives.

---

# Conclusion

* Logistic Regression performs competitively despite simplicity
* XGBoost offers marginal improvements in some areas
* Threshold selection has a major impact on results
* Proper preprocessing and labeling are essential

---



